---
title: "Reproducible Research: Peer Assessment 1"
output: bookdown::html_document2
---


## Loading and preprocessing the data
This section of code sets global options, loads the raw data and processes it
such that steps by interval are aggregated to daily steps.

```{r setoptions,include=FALSE}
knitr::opts_chunk$set(warning=FALSE,message=FALSE,fig.keep="all")
```

```{r readprocdata}
fileUrl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
temp<-tempfile() # create a temporary file
download.file(fileUrl,temp) #download file into temporary location
activity<-read.csv(unz(temp,"activity.csv"))
library(dplyr)
activityGrp<-group_by(activity,date)
activityDaily<-summarise(activityGrp,dailystep=sum(steps,na.rm=TRUE))
source("formatNum.R")
```

## What is mean total number of steps taken per day?
The median number of steps taken each day was 
**`r formatNum(median(activityDaily$dailystep),0)`**, whilst the mean was
**`r formatNum(mean(activityDaily$dailystep),0)`**, when missing values were not
considered. A histogram displays the distribution of total steps in a day below 
.Note that a considerable number of observations had missing 
values, denoted in the 0 bar.

```{r plothist,fig.cap="Histogram of total number of steps taken per day"}
library(ggplot2)
qplot(dailystep,data=activityDaily,ylab="Frequency",xlab="Total steps in a day",
      binwidth=2500,geom="histogram",fill=I("springgreen2"),col=I("springgreen3"))
```


## What is the average daily activity pattern?
```{r groupByInt}
activityGrp<-group_by(activity,interval)
activityInt<-summarise(activityGrp,avgstep=mean(steps,na.rm=TRUE))
```
The maximum number of steps on average was observed on the **`r activityInt$interval[which.max(activityInt$avgstep)]`th** interval.
```{r plotscatt,fig.cap="Time series plot of average number of steps versus interval"}
qplot(interval,avgstep,data=activityInt,ylab="Average number of steps during 
      interval",xlab="Interval",col=I("steelblue1"),geom="line")
```


## Imputing missing values
``` {r missingcalc}
totalrows<-nrow(activity)
completerows<-sum(complete.cases(activity))
missingrows<-totalrows-completerows
library(Hmisc)
imputedsteps<-impute(activity$steps,fun=mean(activity$steps,na.rm=TRUE))
activityNew<-activity
activityNew$steps<-imputedsteps
activityGrp<-group_by(activityNew,date)
activityNewDaily<-summarise(activityGrp,dailystep=sum(steps,na.rm=TRUE))
```
The dataset contained many missing values. Out of 
**`r formatNum(totalrows,0)`** observations, only **`r formatNum(completerows,0)`**
were complete with the remaining **`r formatNum(missingrows,0)`** missing data 
on the number of steps.When missing values were imputed, the median number of 
steps taken each day was **`r formatNum(median(activityNewDaily$dailystep),0)`**,
whilst the mean was **`r formatNum(mean(activityNewDaily$dailystep),0)`**, both
values higher than estimated when excluding missing data.
```{r plothist2,fig.cap="Histogram of total number of steps taken per day with imputed data"}
qplot(dailystep,data=activityNewDaily,ylab="Frequency",xlab="Total steps in a day (imputed)", binwidth=2500,geom="histogram",fill=I("violetred"),col=I("violetred1"))
```


## Are there differences in activity patterns between weekdays and weekends?
The plots below display the average number of steps at different intervals
during weekdays and weekends.
``` {r plotdaytype,fig.cap="Time series plot of interval versus average number of steps by type of day"}
library(lubridate)
activity$date<-ymd(activity$date)
activity$day<-weekdays(activity$date)
activity$dayType<-"Weekday"
activity$dayType[activity$day=="Saturday"]<-"Weekend"
activity$dayType[activity$day=="Sunday"]<-"Weekend"
activityByDayType<-aggregate(activity,by=list(activity$interval,activity$dayType),
                             FUN=mean,na.rm=TRUE)
library(tidyverse)
activityByDayType<-rename(activityByDayType,DayType=Group.2,Interval=Group.1)

qplot(Interval,steps,data=activityByDayType,facets=DayType~.,
ylab="Average number of steps during interval",xlab="Interval",
col=DayType,geom="line")
```


